Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-13 doc/manuscript/translocation.qmd
Convert manuscript to PNAS after small edits

The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
A total of 33 annotated genes were found in the 9 shared regions (SI).
The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had a low $\pi_{diff}$ value <!--# Negative value shown in 2A? -->, consistent with balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot). This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.

@@

Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).

@@

We conducted two analyses.
In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
Group-level effects included site_id or translocation_id.
The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), winter severity in the year following translocation (snow_t1), site elevation (elevation), donor population (donor_72996), sex (sex_male), frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first)).
In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3\--16).
Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
We checked convergence of the MCMC algorithm using trace plots and Rhat.
Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.

@@

We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI for details].
To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
We used an analysis workflow similar to that described above.
Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
All models used a negative binomial family.
Priors, chains, iterations, and checks were as described above.
Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. --> <!--# Add  -->

@@

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to growth or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity into our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.

@@

We thank the following for important contributions to this study: E.
Hegeman and A.
Lindauer--project and data management, field and laboratory work; A.
Barbella and K.
Rose--laboratory work; C.
Kamoroff and numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S. Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National Science Foundation Grants EF-0723563 (to C. Briggs), DEB-1557190 (to C. Briggs), and DBI-2120084 (to C. Richards-Zawacki).

@@

Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
We simulated our model for 16 years, repeating the simulation 50,000 times.
For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.

>>>>

The splined window analysis identified 33 outlier regions for $\pi_{diff}$ and 58 outlier regions for *F~ST~* (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
A total of 33 annotated genes were found in the 9 shared regions (SI).
The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had a low $\pi_{diff}$ value<!--# Negative value shown in 2A? -->, consistent with balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D, @fig-synteny-plot). This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
In *Xenopus*, five of these genes occur in the extended major histocompatibility complex (MHC) Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.

@@

Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where yearly adult survival probability $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis).

@@

We conducted two analyses.
In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
Group-level effects included site_id or translocation_id.
The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), winter severity in the year following translocation (snow_t1), site elevation (elevation), donor population (donor_72996), sex (sex_male), frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first)).
In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3\--16).
Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
We checked convergence of the MCMC algorithm using trace plots and Rhat.
Code to create the datasets and fit models is available from the following GitHub repository: [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).

@@

We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI for details].
To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
We used an analysis workflow similar to that described above.
Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
All models used a negative binomial family.
Priors, chains, iterations, and checks were as described above.
Code to create the dataset and fit models is available from the following GitHub repository: [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).
<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. --> <!--# Add  -->

@@

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values] : SI), we performed four analyses (described in detail in the SI).
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity into our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.

@@

We thank the following for important contributions to this study: E.
Hegeman and A.
Lindauer--project and data management, field and laboratory work; A.
Barbella and K.
Rose--laboratory work; C.
Kamoroff and numerous summer technicians--field work; staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S. Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory--research permits, logistical support, and/or field assistance.
This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.); and National Science Foundation Grants EF-0723563 (to C. Briggs), DEB-1557190 (to C. Briggs), DEB-2133401 (to M.Q.W), and DBI-2120084 (to C. Richards-Zawacki).

@@

Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
We simulated our model for 16 years, repeating the simulation 50,000 times.
For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.  Code to replicate the analyses can be found at [https://github.com/SNARL1/translocation](https://github.com/SNARL1/translocation).
